---
title: "Practicum 01"
output: html_notebook
---

Part 1-1: Create a dataframe with 4 variables. The 4 variables should be doctor_type, 
doctor_lastname, location, AVG_Rating. 

```{r}
doctor_type <- c("PCP", "Psychiatrist", "Surgeon", "Anesthesia")
doctor_lastname <- c("Smith", "Dame", "Jones", "Zayas")
location <- c("MA", "ME", "NH", "VT")
AVG_Rating <- c(7,9,8,9)

medical_df <- data.frame(doctor_type, doctor_lastname, location, AVG_Rating)

print(medical_df)
```

part 1-2: Using the data frame above…

Select row 1 in column 2, what was selected?

1 refers to the observations associated with Dr. Smith, and column 2 refers
to the variables which contain the lastnames for all doctors. Selecting row 1
column 2 returns the last name "Smith". 


Select Rows 2 through 4, what was selected?

Rows 2 through 4 contain all observations associated with Drs Dame, Jones, and
Zayas. this includes type, lastname, location, and average rating. 


Select the last column, what was selected?

Selecting the last column returns all values for the variable "AVG_Rating", which 
is presumably the average rating of each doctor.

```{r}
#Row 1, Column 2
print(medical_df[1,2])

#Rows 2 through 4
print(medical_df[2:4,])

#Last Column
print(medical_df[,4])
```


Part 1-3: Using the dataset Mtcars create a scatterplot showing the relations between any two 
variables of your choosing. Explain why you picked those variables. Each point should also 
be based on a color scheme of your choosing.

The question that I am seeking to answer is how car weight impacts MPG rating. It is imaginable that 
heavier cars would require more energy to move, and the increased energy need would need to come
from an increase in fuel consumption. 

It is also worth inspecting the size of the engine along with the weight of the vehicle, 
as engine size may also be related to fuel consumption.

Data is plotted with weight on the X axis and miles per gallon on the y axis. Each point is colored by
cylinder number, which is representative of the engine size. 

Before drawing conclusions, statistical analysis was performed to remove outliers. Analysis of Z-scores
shows that there are no values three standard deviations from the mean in the data frame for either variable, so no curation of the 
data points occurred. 

Correlation analysis using the Pearson methodology returns a value of -0.8696594. This values rounds into 
the criteria held by research science of ± 0.9, indicating a negative correlation between vehicle weight and
mpg rating. Inspecting the coloration of the points also shows a distinct trend between mpg rating and the 
size of the engine.

```{r}
library(tidyverse)

#Plot variables with Outliers
ggplot(data = mtcars, mapping = aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point() + 
  scale_x_log10() +
  labs(
    title = "Vehicle Weight vs. MPG Rating",
    caption = "Source: ggplot2 mtcars Dataset") +
  xlab("Vehicle Weight") +
  ylab("Miles-per-Gallon (MPG)") +
  theme_bw()

#Calculate statistics
mtcars_df <- mtcars

mean_weight <- mean(mtcars_df$wt)
sd_weight <- sd(mtcars_df$wt)
mtcars_df$z1 <- abs((mean_weight-mtcars_df$wt)/sd_weight)

mean_mpg <- mean(mtcars_df$mpg)
sd_mpg <- sd(mtcars_df$mpg)
mtcars_df$z2 <- abs((mean_mpg-mtcars_df$mpg)/sd_mpg)

#Calculate correlation coefficient
cor(x = mtcars$wt, y = mtcars$mpg, method = "pearson")
```

Part 1-4: Using the dataset MTcars perform a summary statistic on the dataset and calculate the 
pearson coefficient of the correlation R picking two variables of choice. Explain why you 
picked those variables and explain the purpose of a pearson coefficient.

Analysis in part 1-3 suggests that car weight impacts mileage, but it would be an assumption to state that
this relationship is causal. We can verify the claim that 'heavy cars have poor mileage because heavier machines
require more energy to move" by exploring the relationship between car weight and gross horsepower. This will rule 
out alternative explanations, such as "heavy cars have poor gas mileage because they are built inefficiently". 

First, summary statistics for the entire data set are produced by the code below.

Before evaluation, statistical analysis was performed to remove outliers. Analysis of Z-scores
shows that there are no values three standard deviations from the mean in the data frame for either variable, 
so no curation of the data points occurred before calculating the correlation coefficient. 

Correlation analysis using the Pearson methodology returns a value of 0.6587479. This value indicate a non-significant, 
positive relationship between these two variables. This value is in essence an aggregate measurement of how much each 
data point deviates from regression line modeled during the comparison of the two variables.

Overall, these results suggests that the need for higher horsepower may not necessarily be the reason that heavier cars
tend to have lower gas mileage.

```{r}
#Summary statistics
summary(mtcars)

#Analyze data set for outliers within either variable
mtcars_df2 <- mtcars
mean_weight <- mean(mtcars_df2$wt)
sd_weight <- sd(mtcars_df2$wt)
mtcars_df2$z1 <- abs((mean_weight-mtcars_df2$wt)/sd_weight)

mean_hp <- mean(mtcars_df2$hp)
sd_hp <- sd(mtcars_df2$hp)
mtcars_df2$z2 <- abs((mean_hp-mtcars_df2$hp)/sd_hp)

summary(mtcars_df2)

#Perform a pearson coefficient test
cor(x = mtcars$wt, y = mtcars$hp, method = "pearson")
```

Part 2: You are given the task of performing a comprehensive analysis of the admission statistics from 2007 to 
2019 and summarize your findings with an accompanying narrative that explains your process-flow.

part 2-1: Load the data, directly from the URL, into your R environment

The code below loads the .csv file into RStudio as a data frame.

```{r}
#Data is imported from Data.NY.gov 
addiction_df <- read.csv("https://data.ny.gov/api/views/ngbt-9rwf/rows.csv")

```

part 2-2: Evaluate the dataset to determine what data preparation steps are needed and perform them. 
At a minimum, ensure that you discuss the distribution of the data, outliers and prepare any helpful
summary statistics to support your analysis.

- Turn age into leveled factors.

Outliers of admissions data evaluated using a z-score criteria of equal to or greater than 3. The filtered outliers were evaluated in Figure 1 and it was observed that the majority of the outlying values came from urban counties within the city. Since artificial variance could occur from comparing rural low population counties to densely population urban counties, outlier analysis was re-performed with the dataset divided between counties inside and outside of NYC. Since there is no objective measurement of population per county available 


#Analyze admissions individually by substance. Looking at overall admissions is of course going to be skewed.
#Filter for admissions statistics from 2020. Only should evaluate from 2007 - 2019.
#None is under primary substance group. Maybe lets remove it as a missing value.

```{r}
library(lubridate)
#Preparation of data
addiction_clean <- addiction_df %>%
  mutate(across(c(Age.Group, Program.Category, Service.Type, Primary.Substance.Group), as.factor)) %>%
  filter(Year <= 2019) %>%
  mutate(Year = as.Date(paste(Year, 1, 1, sep = "-")))

levels(addiction_clean$Age.Group) <- c("Under 18", "18 thru 24", "25 thru 34", "35 thru 44", "45 thru 54", "55 and Older") 
#-------------------------------------------------------------------------------
summary(addiction_clean$Admissions)

ggplot(data = addiction_clean, mapping = aes(x = Admissions)) +
  geom_histogram(binwidth = 10, fill = "cadetblue") +
  labs(
    title = "Distribution of the Admissions statistics",
    caption = "Source: OASAS Chemical Dependance Treatment Program"
  ) +
  xlab("Admissions") + 
  ylab("Count") +
  theme_bw()

mean_admissions <- mean(addiction_clean$Admissions)
sd_admissions <- sd(addiction_clean$Admissions)
addiction_clean$z <- abs((mean_admissions-addiction_clean$Admissions)/sd_admissions)
addiction_outlier <- filter(addiction_clean, z >= 3)
addiction_clean <- filter(addiction_clean, z < 3)

summary(addiction_clean$Admissions)

ggplot(data = addiction_clean, mapping = aes(x = Admissions)) +
  geom_histogram(binwidth = 10, fill = "cadetblue") +
  labs(
    title = "Distribution of the Admissions statistics",
    caption = "Source: OASAS Chemical Dependance Treatment Program"
  ) +
  xlab("Admissions") + 
  ylab("Count") +
  theme_bw()

```

Part 2-3: Structure the data relationally, at a minimum, you should have four tibbles or data frames as 
follows: County, Program_category, Primary_Substance_Group, Admissions Data.

Within the county data frame county, the boroughs of NYC are specified by their county code 'NY' and the borough name in parentheses in order to create a distinctive key:value relationship. County codes and names are associated as a data frame, as well as program codes and names. These two data frames essentially serve as primary key dictionaries for the altered dataset to use. 

The dataset is restructured through use of these primary keys, by using the key-value relationship in these newly defined data frames as patterns for replacement in the cleaned data frame. By using this technique, county names were able to be replaced with county codes, program names with program codes, and substance names with substance codes. With these codes in place, relational connections can be established and it large environments of multiple data frames can be constructed.

```{r}
library(stringi)
#-------------------------------------------------------------------------------
#Create County data frame.

county_code <- c("AL", "AG", "NY (Bronx)", "BM", "CA", "CY", "CH", "CM",
                 "CN", "CL", "CO", "CR", "DE", "DU", "ER", "ES",
                 "FR", "FU", "GE", "GR", "HE", "JE", "NY (Kings)",
                 "LE", "LI", "MA", "MO", "MG", "NA", "NY (New York City)", "NI",
                 "ON", "OD", "OT", "OR", "OL", "OS", "OG", "PU",
                 "NY (Queens)", "RE", "NY (Richmond)", "RO", "SL", "SA", "SC", "SH",
                 "SY", "SE", "ST", "SU", "SV", "TI", "TO", "UL",
                 "WR", "WS", "WA", "WE", "WY", "YA"
                 )

County.of.Program.Location <- c(addiction_clean$County.of.Program.Location)
County.of.Program.Location <- unique(County.of.Program.Location)
county <- data.frame(county_code, County.of.Program.Location)
#-------------------------------------------------------------------------------
#Create program category data frame.

program_code <- c("CR", "IN", "OT", "OP", "RE", "SP")
program_name <- unique(c(addiction_clean$Program.Category))

program_category <- data.frame(program_code, program_name)
#-------------------------------------------------------------------------------
#Primary substance group data frame.
substance_code <- c("A", "C", "H", "M", "O", "AO", "NA")
substance_name <- c("Alcohol", "Cocaine incl Crack", "Heroin", "Marijuana incl Hashish", "Other Opioids", "All Others", "None")
primary_substance_group <- data.frame(substance_code, substance_name)
#-------------------------------------------------------------------------------
#Admissions data restructured relationally.

addiction_relational <- addiction_clean %>%
  mutate(Primary.Substance.Group = stri_replace_all_regex(Primary.Substance.Group, 
                                                          pattern = primary_substance_group$substance_name,
                                                          replacement = primary_substance_group$substance_code,
                                                          vectorize_all = FALSE)) %>%
                            rename("substance_code" = Primary.Substance.Group) %>%
  
  mutate(Program.Category = stri_replace_all_regex(Program.Category, 
                                                          pattern = program_category$program_name,
                                                          replacement = program_category$program_code,
                                                          vectorize_all = FALSE)) %>%
                            rename("program_code" = Program.Category) %>%

  mutate(County.of.Program.Location = stri_replace_all_regex(County.of.Program.Location, 
                                                          pattern = county$County.of.Program.Location,
                                                          replacement = county$county_code,
                                                          vectorize_all = FALSE)) %>%
                            rename("county_code" = County.of.Program.Location)
```



part 2-4: Create a function called annualAdmissions() that derives the total number of reported 
admissions that transpired each year, for the entire state of NY and displays the results using a line 
chart. Annotate the chart to show the year with the highest number of admissions. Note: the year 
should be on the x-axis and the number of admissions on the y-axis. Explain the chart

Figure 1 below demonstrates the total admittance to the OASAS Chemical Dependence Treatment Program per year between the years 2007-2019. The first notable rise in treatment center admittance was observed in 2008, which coincides with a US economic recession which specifically targeted the housing market. Examining the breakdown of admittance by county in 2008, rural counties where housing is owned rather than rented such as Albany showed overall admittance on-par with densely populated urban counties within NYC as portions of the population invested in real estate found themselves in poverty. Hard economic conditions are known to promote drug abuse, so the peak in 2008 can be considered consistent.

The most significant increase in admittance can be observed in 2016, where the total number sharply increases. 2016 is commonly known as the start of the 3rd wave of the opioid epidemic, when the re-emergence of synthetic opioids such as Fentanyl led many users to turn to heroin as a cheaper and more available option. The impact of this can be found in Figure 3, which shows a breakdown of abuse by substance for the year 2016. In this figure, it can be seen that the most common substance that a patient sought treatment for was heroin in 2016.



#Add annotations to this graph
```{r}
#First assume for 2007.

annualAdmissions <- function() {
addiction_annual <- addiction_clean %>%
  group_by(Year) %>%
  summarise(sum(Admissions))

ggplot(data = addiction_annual, mapping = aes(x = Year, y = `sum(Admissions)`)) +
  geom_line(color = "red") +
  scale_y_log10() +
  labs(
    title = "Figure 1: Admisssions per Year",
    caption = "Source: OASAS Chemical Dependance Treatment Program"
  ) +
  xlab("Year") +
  ylab("Admissions per year") +
  annotate(geom="text", x=as.Date("2008-01-01"), y=191000,
           label="Economic Recession") +
   annotate(geom="text", x=as.Date("2016-01-01"), y=195000,
           label="Resurgance of Synthetic Opioids") +
  theme_bw()
}
annualAdmissions()
#-------------------------------------------------------------------------------
#2008 Recession
recession_admissions <- addiction_clean %>%
  filter(Year == "2008-01-01") %>%
  group_by(County.of.Program.Location) %>%
  summarise(sum(Admissions))

ggplot(data = recession_admissions, mapping = aes(reorder(x = County.of.Program.Location, desc(`sum(Admissions)`)), y = `sum(Admissions)`, fill = `sum(Admissions)`)) +
  geom_bar(stat = "identity") +
  scale_y_log10() +
  labs(
    title = "Figure 2: Admisssions per County in 2008",
    caption = "Source: OASAS Chemical Dependance Treatment Program"
  ) +
  xlab("County") +
  ylab("Admissions per year") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
#-------------------------------------------------------------------------------
#2016 Opioid Epidemic
substance_admissions <- addiction_clean %>%
  filter(Year == "2016-01-01") %>%
  group_by(Primary.Substance.Group) %>%
  summarise(sum(Admissions))

ggplot(data = substance_admissions, mapping = aes(reorder(x = Primary.Substance.Group, desc(`sum(Admissions)`)), y = `sum(Admissions)`, fill = "dodgerblue")) +
geom_bar(stat = "identity") +
   labs(
    title = "Figure 3: Admisssions per Substance in 2016",
    caption = "Source: OASAS Chemical Dependance Treatment Program"
  ) +
  xlab("Primary Substance Group") +
  ylab("Admissions per Substance in 2016") +
  theme_bw() +
  theme(axis.text.x.bottom = element_text(size = 7))
  #theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```


Part 2-5: Analyze the percentage of admissions for each county and visualize the results for the top 5 
counties using a bar chart. Explain the results. Note: ensure that you join any related dataframes/ 
tibbles.

```{r}
addiction_county <- addiction_relational %>%
  full_join(county, by = "county_code") %>%
    group_by(County.of.Program.Location) %>%
      summarise(sum(Admissions)) %>%
        rename("county_admissions" = `sum(Admissions)`) %>%
      mutate(percent = (county_admissions / sum(county_admissions)) * 100) %>%
    arrange(desc(percent))

addiction_county_top <- head(addiction_county, n = 5L)


ggplot(data = addiction_county_top, mapping = aes(x = reorder(County.of.Program.Location, desc(percent)), y = percent)) + 
  geom_bar(position = "dodge", stat = "identity", fill = "deepskyblue2") +
  labs(
    title = "Top 5 NY Counties for Drug Rehab Admittance, by percentage",
    Caption = "Source: OASAS Chemical Dependance Treatment Program"
  ) +
  xlab("NY County") +
  ylab("Overall % of Rehab Admissions in NY State") +
  theme_bw()
print(addiction_county)
```


Part 2-6: Filter the data, using a regular expression, and extract all admissions to the various “Rehab” 
facilities; i.e. your regex should match all facilities that include the word rehab, rehabilitation, etc. 
Using the filtered data, identify which substance is the most prominent among each age group. 
Visualize and explain the results

The cleaned dataset below is filtered for keywords regarding rehabilitation clinics. Data is grouped by age group and substance used. Admissions data regarding each substance for each age grouped was summed and the highest represented substance per age group was extracted and plotted below. 

The overall trend observed is that younger people most commonly enter rehabilitation clinics for substance abuse treament regarding heroin, while older individuals commonly enter rehabilitation clinics for alcohol abuse treatment. Since this data was collected in NYC, there could be potential bias in that alcohol is more freely available to some age groups versus others. This can be seen in Figure 2, which displays the use of each substance by age group. Alcohol abuse rates among those 18 and under are low when compared to Marijuana and Heroin, and then abuse rates sharply rise in the age group where alcohol becomes legal to purchase.

the high heroin rates among young people may also be a product of some natural bias. Examining the age breakdown of each substance in Figure 2 shows that heroin abuse seems to peak at age 18-24 and sharply decline with age. Heroin is a notoriously dangerous drug which can easily and quickly trigger lethal health events. It is possible that abuse statistics regarding the drug are skewed in this manner due to the abuser dying before treatment can be received.

```{r}
addiction_filter <- addiction_clean %>%
  filter(str_detect(Service.Type, "Rehab|Rehabilitation|Rehab$")) %>%
    group_by(Age.Group, Primary.Substance.Group) %>%
      summarise(sum(Admissions)) %>%
        summarise(max(`sum(Admissions)`), across()) %>%
          rename("max_admissions" = `max(\`sum(Admissions)\`)`) %>%
      rename("Admissions" = `sum(Admissions)`) %>%
    filter(max_admissions == Admissions) %>%
  select(-max_admissions)

print(addiction_filter)

addiction_stats <- addiction_clean %>%
  filter(str_detect(Service.Type, "Rehab|Rehabilitation|Rehab$")) %>%
    group_by(Age.Group, Primary.Substance.Group) %>%
      summarise(sum(Admissions))

#-------------------------------------------------------------------------------

ggplot(data = addiction_filter, mapping = aes(x = Age.Group, y = Admissions, fill = Primary.Substance.Group)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Figure 1: Most Commonly Abused Substance per Age Group by Admittance",
    caption = "Source: OASAS Chemical Dependance Treatment Program"
  ) +
  xlab("Age Group") +
  ylab("Rehab Clinic Admittance") +
  theme_bw()

ggplot(data = addiction_stats, mapping = aes(x = Primary.Substance.Group, y = `sum(Admissions)`, fill = Age.Group)) +
  geom_bar(position = "dodge", stat = "Identity") +
  labs(
    title = "Figure 2: Abuse rate of substances per Age Group",
    caption = "Source: OASAS Chemical Dependance Treatment Program"
  ) +
  xlab("Age Group") +
  ylab("Rehab Clinic Admittance") +
  theme_bw() +
  theme(axis.text.x.bottom = element_text(size = 7))
  #(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```