---
title: "Practicum 01"
output: html_notebook
---

Part 1-1: Create a dataframe with 4 variables. The 4 variables should be doctor_type, 
doctor_lastname, location, AVG_Rating. 

```{r}
doctor_type <- c("PCP", "Psychiatrist", "Surgeon", "Anesthesia")
doctor_lastname <- c("Smith", "Dame", "Jones", "Zayas")
location <- c("MA", "ME", "NH", "VT")
AVG_Rating <- c(7,9,8,9)

medical_df <- data.frame(doctor_type, doctor_lastname, location, AVG_Rating)

print(medical_df)
```

part 1-2: Using the data frame above…

Select row 1 in column 2, what was selected?

1 refers to the observations associated with Dr. Smith, and column 2 refers
to the variables which contain the lastnames for all doctors. Selecting row 1
column 2 returns the last name "Smith". 


Select Rows 2 through 4, what was selected?

Rows 2 through 4 contain all observations associated with Drs Dame, Jones, and
Zayas. this includes type, lastname, location, and average rating. 


Select the last column, what was selected?

Selecting the last column returns all values for the variable "AVG_Rating", which 
is presumably the average rating of each doctor.

```{r}
#Row 1, Column 2
print(medical_df[1,2])

#Rows 2 through 4
print(medical_df[2:4,])

#Last Column
print(medical_df[,4])
```


Part 1-3: Using the dataset Mtcars create a scatterplot showing the relations between any two 
variables of your choosing. Explain why you picked those variables. Each point should also 
be based on a color scheme of your choosing.

The question that I am seeking to answer is how car weight impacts MPG rating. It is imaginable that 
heavier cars would require more energy to move, and the increased energy need would need to come
from an increase in fuel consumption. 

It is also worth inspecting the size of the engine along with the weight of the vehicle, 
as engine size may also be related to fuel consumption.

Data is plotted with weight on the X axis and miles per gallon on the y axis. Each point is colored by
cylinder number, which is representative of the engine size. 

Before drawing conclusions, statistical analysis was performed to remove outliers. Analysis of Z-scores
shows that there are no values three standard deviations from the mean in the data frame for either variable, so no curation of the 
data points occurred. 

Correlation analysis using the Pearson methodology returns a value of -0.8696594. This values rounds into 
the criteria held by research science of ± 0.9, indicating a negative correlation between vehicle weight and
mpg rating. Inspecting the coloration of the points also shows a distinct trend between mpg rating and the 
size of the engine.

```{r}
library(tidyverse)

#Plot variables with Outliers
ggplot(data = mtcars, mapping = aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point() + 
  scale_x_log10() +
  labs(
    title = "Vehicle Weight vs. MPG Rating",
    caption = "Source: ggplot2 mtcars Dataset") +
  xlab("Vehicle Weight") +
  ylab("Miles-per-Gallon (MPG)") +
  theme_bw()

#Calculate statistics
mtcars_df <- mtcars

mean_weight <- mean(mtcars_df$wt)
sd_weight <- sd(mtcars_df$wt)
mtcars_df$z1 <- abs((mean_weight-mtcars_df$wt)/sd_weight)

mean_mpg <- mean(mtcars_df$mpg)
sd_mpg <- sd(mtcars_df$mpg)
mtcars_df$z2 <- abs((mean_mpg-mtcars_df$mpg)/sd_mpg)

#Calculate correlation coefficient
cor(x = mtcars$wt, y = mtcars$mpg, method = "pearson")
```

Part 1-4: Using the dataset MTcars perform a summary statistic on the dataset and calculate the 
pearson coefficient of the correlation R picking two variables of choice. Explain why you 
picked those variables and explain the purpose of a pearson coefficient.

Analysis in part 1-3 suggests that car weight impacts mileage, but it would be an assumption to state that
this relationship is causal. We can verify the claim that 'heavy cars have poor mileage because heavier machines
require more energy to move" by exploring the relationship between car weight and gross horsepower. This will rule 
out alternative explanations, such as "heavy cars have poor gas mileage because they are built inefficiently". 

First, summary statistics for the entire data set are produced by the code below.

Before evaluation, statistical analysis was performed to remove outliers. Analysis of Z-scores
shows that there are no values three standard deviations from the mean in the data frame for either variable, 
so no curation of the data points occurred before calculating the correlation coefficient. 

Correlation analysis using the Pearson methodology returns a value of 0.6587479. This value indicate a non-significant, 
positive relationship between these two variables. This value is in essence an aggregate measurement of how much each 
data point deviates from regression line modeled during the comparison of the two variables.

Overall, these results suggests that the need for higher horsepower may not necessarily be the reason that heavier cars
tend to have lower gas mileage.

```{r}
#Summary statistics
summary(mtcars)

#Analyze data set for outliers within either variable
mtcars_df2 <- mtcars
mean_weight <- mean(mtcars_df2$wt)
sd_weight <- sd(mtcars_df2$wt)
mtcars_df2$z1 <- abs((mean_weight-mtcars_df2$wt)/sd_weight)

mean_hp <- mean(mtcars_df2$hp)
sd_hp <- sd(mtcars_df2$hp)
mtcars_df2$z2 <- abs((mean_hp-mtcars_df2$hp)/sd_hp)

summary(mtcars_df2)

#Perform a pearson coefficient test
cor(x = mtcars$wt, y = mtcars$hp, method = "pearson")
```









