---
title: "Practicum 3"
output: html_notebook
---


Main objective: Determine what factors contribute to tip amount for NYC taxi drivers.

#All libraries are allowed.

Question 1: 

Part 1: • Load the NYC Green Taxi Trip Records data into a data frame or tibble

```{r}
library(tidyverse)
library(caret)
library(psych)

tripdata.df <- read.csv("2018_Green_Taxi_Trip_Data.csv", header = TRUE)

```

Part 2: • explore the data to identify any patterns and analyze the relationships between the 
features and the target variable i.e. tip amount. At a minimum, you should analyze: 1) the distribution,
2) the correlations 3) missing values and 4) outliers — provide supporting visualizations and explain 
all your steps.

Initial summary statistics are printed below. NA values are counted and a proportion of how many values are missing is calculated. The 
column 'ehail fee' is completely missing, and was removed below. The column "Trip type" is largely in-tact with only three missing
values present, and missing values will be replaced with the most common type of trip (1).

#Use pickup and dropoff time to mutate a new column called triplength
#Since there is only one value for trip_type, it will not be useful in the correlation matrix and has been removed.
#Things strongly correlated to tip amount seem to be the type of payment, the total amount of the ride, and the distance of the trip. 
#Pickup and dropoff locations have a pretty binomial distribution.
#Check for outliers: Passenger_count, trip_distance, fare_amount, tip_amount, tolls_amount, total_amount,

```{r}
#Summary Statistics
glimpse(tripdata.df)
dim(tripdata.df)
summary(tripdata.df)

#Count NA values
for(i in 1:ncol(tripdata.df)){
  print(colnames(tripdata.df[i]))
  print(sum(is.na(tripdata.df[,i])))
  print(as.double(sum(is.na(tripdata.df[,i])) / nrow(tripdata.df)*100))
}

#Handling Missing Values
tripdata.clean <- tripdata.df %>%
  select(-ehail_fee)
tripdata.clean$trip_type <- replace_na(1)


#Conversion of variables to numerics for use in model.
tripdata.clean <- tripdata.clean %>%
  mutate(fare_amount = gsub(",", "", fare_amount)) %>%
    mutate(total_amount = gsub(",", "", total_amount)) %>%
      mutate(fare_amount = as.numeric(fare_amount)) %>%
        mutate(total_amount = as.numeric(total_amount)) %>%
      mutate(lpep_pickup_datetime = as.POSIXct(lpep_pickup_datetime, format = "%m/%d/%Y %H:%M")) %>%
    mutate(lpep_dropoff_datetime = as.POSIXct(lpep_dropoff_datetime, format = "%m/%d/%Y %H:%M")) %>%
  mutate(trip_time = lpep_dropoff_datetime - lpep_pickup_datetime)
  
#Conversion of categorical variables to numeric dummy variables
tripdata.clean$store_and_fwd_flag <- ifelse(tripdata.clean$store_and_fwd_flag == "N", 0, 1)


#Distribution before outlier removal. For the love of god turn this into a function. 
#Argument for variable name and argument for title / xlab.

ggplot(data = tripdata.clean, mapping = aes(x = passenger_count)) +
  geom_histogram(binwidth = 1, color = "black", fill = "lightblue") +
  labs(title = "Pre-Removal distribution of Passenger Count",
       caption = "2018 NYC Traffic data") +
  theme_bw()

ggplot(data = tripdata.clean, mapping = aes(x = trip_distance)) +
  geom_histogram(binwidth = 1, color = "black", fill = "lightblue") +
  labs(title = "Pre-Removal Distribution of Trip Distance",
       caption = "2018 NYC Traffic data") +
  theme_bw()

ggplot(data = tripdata.clean, mapping = aes(x = fare_amount)) +
  geom_histogram(binwidth = 5, color = "black", fill = "lightblue") +
  labs(title = "Pre-Removal Distribution of Fare Amount",
       caption = "2018 NYC Traffic data") +
  theme_bw()

ggplot(data = tripdata.clean, mapping = aes(x = tip_amount)) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "lightblue") +
  labs(title = "Pre-Removal Distribution of Tip Amount",
       caption = "2018 NYC Traffic data") +
  theme_bw()

ggplot(data = tripdata.clean, mapping = aes(x = tolls_amount)) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "lightblue") +
  labs(title = "Pre-Removal Distribution of Tolls Amount",
       caption = "2018 NYC Traffic data") +
  theme_bw()

ggplot(data = tripdata.clean, mapping = aes(x = total_amount)) +
  geom_histogram(binwidth = 5, color = "black", fill = "lightblue") +
  labs(title = "Pre-Removal Distribution of Total Amount",
       caption = "2018 NYC Traffic data") +
  theme_bw()


#Outlier removal
passenger.mean <- mean(tripdata.clean$passenger_count)
passenger.sd <- sd(tripdata.clean$passenger_count)
tripdata.clean$zpassenger_count <- abs((passenger.mean - tripdata.clean$passenger_count)/passenger.sd)

distance.mean <- mean(tripdata.clean$trip_distance)
distance.sd <- sd(tripdata.clean$trip_distance)
tripdata.clean$ztrip_distance <- abs((distance.mean - tripdata.clean$trip_distance)/distance.sd)

fare.mean <- mean(tripdata.clean$fare_amount)
fare.sd <- sd(tripdata.clean$fare_amount)
tripdata.clean$zfare_amount <- abs((fare.mean - tripdata.clean$fare_amount)/fare.sd)

tip.mean <- mean(tripdata.clean$tip_amount)
tip.sd <- sd(tripdata.clean$tip_amount)
tripdata.clean$ztip_amount <- abs((tip.mean - tripdata.clean$tip_amount)/tip.sd)

tolls.mean <- mean(tripdata.clean$tolls_amount)
tolls.sd <- sd(tripdata.clean$tolls_amount)
tripdata.clean$ztolls_amount <- abs((tolls.mean - tripdata.clean$tolls_amount)/tolls.sd)

total.mean <- mean(tripdata.clean$total_amount)
total.sd <- sd(tripdata.clean$total_amount)
tripdata.clean$ztotal_amount <- abs((total.mean - tripdata.clean$total_amount)/total.sd)

#This cuts out a ton of data. Bring up to group other ways of handling outliers. 

tripdata.clean <- tripdata.clean %>%
  filter(zpassenger_count < 3) %>%
    filter(ztrip_distance < 3) %>%
      filter(zfare_amount < 3) %>% 
        filter(ztip_amount < 3) %>%
      filter(ztolls_amount < 3) %>%
    filter(ztotal_amount < 3) %>%
  select(!c(zpassenger_count, ztrip_distance, zfare_amount,
            ztip_amount, ztolls_amount, ztotal_amount))


ggplot(data = tripdata.clean, mapping = aes(x = passenger_count)) +
  geom_histogram(binwidth = 1, color = "black", fill = "lightblue") +
  labs(title = "Post-Removal distribution of Passenger Count",
       caption = "2018 NYC Traffic data") +
  theme_bw()

ggplot(data = tripdata.clean, mapping = aes(x = trip_distance)) +
  geom_histogram(binwidth = 1, color = "black", fill = "lightblue") +
  labs(title = "Post-Removal Distribution of Trip Distance",
       caption = "2018 NYC Traffic data") +
  theme_bw()

ggplot(data = tripdata.clean, mapping = aes(x = fare_amount)) +
  geom_histogram(binwidth = 1, color = "black", fill = "lightblue") +
  labs(title = "Post-Removal Distribution of Fare Amount",
       caption = "2018 NYC Traffic data") +
  theme_bw()

ggplot(data = tripdata.clean, mapping = aes(x = tip_amount)) +
  geom_histogram(binwidth = 1, color = "black", fill = "lightblue") +
  labs(title = "Post-Removal Distribution of Tip Amount",
       caption = "2018 NYC Traffic data") +
  theme_bw()

ggplot(data = tripdata.clean, mapping = aes(x = tolls_amount)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "lightblue") +
  labs(title = "Post-Removal Distribution of Tolls Amount",
       caption = "2018 NYC Traffic data") +
  scale_y_log10() +
  theme_bw()

ggplot(data = tripdata.clean, mapping = aes(x = total_amount)) +
  geom_histogram(binwidth = 1, color = "black", fill = "lightblue") +
  labs(title = "Post-Removal Distribution of Total Amount",
       caption = "2018 NYC Traffic data") +
  theme_bw()

#Identify correlations between variables 
tripdata.cor <- tripdata.clean %>%
  select(-lpep_pickup_datetime, -lpep_dropoff_datetime, -trip_time, -trip_type)

correlation.matrix <- cor(tripdata.cor)
print(correlation.matrix)

```



























